\chapter{Fields}

\cmd{FIELD} defines generic tabular data. Fields may be defined at root level or inside a loadcase block, depending on their intended use.

\begin{syntax}
*FIELD, NAME=<name>, DOMAIN=NODE|ELEMENT|IP, COLS=<n>, FILL=ZERO|NAN
<id>, <v1>, <v2>, ...
\end{syntax}

\key{COLS} is required and must be greater than zero. FEMaster limits fields to 64 components. \key{FILL} defaults to \val{ZERO}; \val{NAN} initializes unspecified values with NaN.

\begin{longtable}{@{}p{0.18\linewidth}p{0.72\linewidth}@{}}
\toprule
\textbf{Domain} & \textbf{Meaning} \\
\midrule
\endhead
\val{NODE} & One row per node ID. Used for temperature fields, initial velocities, or prescribed nodal data. \\
\val{ELEMENT} & One row per element ID. Used for topology density and orientation fields. \\
\val{IP} & Integration-point field. Used for data that is tied to element integration points. \\
\bottomrule
\end{longtable}

Data lines may omit individual values; omitted values do not overwrite existing entries. The tokens \val{NAN}, \val{INF}, \val{+INF}, and \val{-INF} are accepted.

\section{Typical references}

\begin{longtable}{@{}p{0.25\linewidth}p{0.25\linewidth}p{0.4\linewidth}@{}}
\toprule
\textbf{Use} & \textbf{Domain/Cols} & \textbf{Referencing command} \\
\midrule
\endhead
Temperature load & \val{NODE}, 1 & \cmd{TLOAD} with \key{TEMPERATUREFIELD}. \\
Initial velocity & \val{NODE}, 6 & \cmd{INITIALVELOCITY}. \\
Topology density & \val{ELEMENT}, 1 & \cmd{TOPODENSITY}. \\
Topology orientation & \val{ELEMENT}, 3 & \cmd{TOPOORIENT}. \\
\bottomrule
\end{longtable}
